Concurrent Programming in Harmony

### 2021-01-24 book/paper.tex
Starting to update the book and remove semicolons

Like Python, Harmony is an imperative,
dynamically typed, and garbage collected programming language.
There are also some important differences:
\begin{itemize}
\item Every statement in Harmony must be terminated by a semicolon
(even \textbf{def} statements).
In Python semicolons are optional and rarely used.
There is no syntactic significance to indentation in Harmony.
Correct indentation in Harmony is encouraged but not enforced.
\item Harmony only supports basic operator precedence or associativity.
Use parentheses liberally to remove ambiguity.
\item Harmony does not (currently) support floating point, iterators, or I/O;
Harmony does support \textbf{for} loops and various ``comprehensions.''
\item Python is object-oriented, supporting classes with methods and
inheritance; Harmony has objects but does not support classes.  Harmony supports
pointers to objects and methods.
\end{itemize}
There are also less important ones that you will discover as
you get more familiar with programming in Harmony.

@@

\begin{code}
\textbf{assert} (\textit{count} == 1) or (\textit{count} == 2);
\end{code}

@@

\begin{figure}
\begin{code}
\begin{verbatim}
   0 Frame __init__ ()
code/Up.hny:1 count = 0;
   1 Push 0
   2 Store count
code/Up.hny:2 done = [ False, False ];
   3 Push [False, False]
   4 Store done
code/Up.hny:4 def incrementer(self):
   5 Jump 32
   6 Frame incrementer self
code/Up.hny:5     count = count + 1;
   7 Load count
   8 Push 1
   9 2-ary +
   10 Store count
code/Up.hny:6     done[self] = True;
   11 Push ?done
   12 LoadVar self
   13 Address
   14 Push True
   15 Store
code/Up.hny:7     await done[1 - self];
   16 Push 1
   17 LoadVar self
   18 2-ary -
   19 Load done
   20 Apply
   21 JumpCond False 16
code/Up.hny:8     assert count == 2, count;
   22 ReadonlyInc
   23 AtomicInc
   24 Load count
   25 Push 2
   26 2-ary ==
   27 Load count
   28 Assert2
   29 AtomicDec
   30 ReadonlyDec
   31 Return
\end{verbatim}
\end{code}
\caption{The first part of the HVM bytecode corresponding to \autoref{fig:inc}.}
\label{fig:inccode}
\end{figure}

@@

However, mutual exclusion by itself is easy to ensure.
For example, we could insert the following code to enter the
critical section:
\begin{code}
\texttt{await False;}
\end{code}
This code will surely prevent two or more threads from being
at label \texttt{@cs} at the same time.
But it does so by preventing \emph{any} thread from reaching
the critical section.
We clearly need another property besides mutual exclusion.

@@

\section*{Exercises}
\begin{problems}
\item Run \autoref{fig:cs} using Harmony.  As there is no protection of the critical
section, mutual exclusion is violated,
the assertion should fail, and a trace should be reported.
Now insert
\begin{code}
\texttt{await False;}
\end{code}
just before entering the critical section
in \autoref{fig:cs} and run Harmony again.
Mutual exclusion is guaranteed but progress is violated.
Harmony should print a trace
to a state from which a terminating state cannot be reached.
Describe in English the difference in the failure reports before
and after inserting the code.
\item See if you can come up with some different approaches that satisfy both
mutual exclusion and progress.  Try them with Harmony and see if they work or not.
If they don't, try to understand why.
Do not despair if you can't figure out how to develop a solution that satisfies both
mutual exclusion and progress---as
we will find out, it is possible but not easy.
\end{problems}

>>>>

Like Python, Harmony is an imperative,
dynamically typed, and garbage collected programming language.
There are also some important differences:
\begin{itemize}
\item Harmony only supports basic operator precedence or associativity.
Use parentheses liberally to remove ambiguity.
\item Harmony does not (currently) support floating point, iterators, or I/O;
Harmony does support \textbf{for} loops and various ``comprehensions.''
\item Python is object-oriented, supporting classes with methods and
inheritance; Harmony has objects but does not support classes.  Harmony supports
pointers to objects and methods.
\end{itemize}
There are also less important ones that you will discover as
you get more familiar with programming in Harmony.

@@

\begin{code}
\textbf{assert} (\textit{count} == 1) or (\textit{count} == 2)
\end{code}

@@

\begin{figure}
\begin{code}
\begin{verbatim}
   0 Frame __init__ ()
code/Up.hny:1 count = 0
   1 Push 0
   2 Store count
code/Up.hny:2 done = [ False, False ]
   3 Push [False, False]
   4 Store done
code/Up.hny:4 def incrementer(self):
   5 Jump 32
   6 Frame incrementer self
code/Up.hny:5     count = count + 1
   7 Load count
   8 Push 1
   9 2-ary +
   10 Store count
code/Up.hny:6     done[self] = True
   11 Push ?done
   12 LoadVar self
   13 Address
   14 Push True
   15 Store
code/Up.hny:7     await done[1 - self]
   16 Push 1
   17 LoadVar self
   18 2-ary -
   19 Load done
   20 Apply
   21 JumpCond False 16
code/Up.hny:8     assert count == 2
   22 ReadonlyInc
   23 AtomicInc
   24 Load count
   25 Push 2
   26 2-ary ==
   27 Load count
   28 Assert2
   29 AtomicDec
   30 ReadonlyDec
   31 Return
\end{verbatim}
\end{code}
\caption{The first part of the HVM bytecode corresponding to \autoref{fig:inc}.}
\label{fig:inccode}
\end{figure}

@@

However, mutual exclusion by itself is easy to ensure.
For example, we could insert the following code to enter the
critical section:
\begin{code}
\texttt{await False}
\end{code}
This code will surely prevent two or more threads from being
at label \texttt{@cs} at the same time.
But it does so by preventing \emph{any} thread from reaching
the critical section.
We clearly need another property besides mutual exclusion.

@@

\section*{Exercises}
\begin{problems}
\item Run \autoref{fig:cs} using Harmony.  As there is no protection of the critical
section, mutual exclusion is violated,
the assertion should fail, and a trace should be reported.
Now insert
\begin{code}
\texttt{await False}
\end{code}
just before entering the critical section
in \autoref{fig:cs} and run Harmony again.
Mutual exclusion is guaranteed but progress is violated.
Harmony should print a trace
to a state from which a terminating state cannot be reached.
Describe in English the difference in the failure reports before
and after inserting the code.
\item See if you can come up with some different approaches that satisfy both
mutual exclusion and progress.  Try them with Harmony and see if they work or not.
If they don't, try to understand why.
Do not despair if you can't figure out how to develop a solution that satisfies both
mutual exclusion and progress---as
we will find out, it is possible but not easy.
\end{problems}
